|

### 2024-09-04 manuscript/paperCMSACversion.Rmd
country of players
edits (and title change)

---
title: |
  |  \large Enforcers are fine, probably.  
author: | 
abstract: |
 | A recent study looked at the relationship between long-term mortality and being an enforcer in the National Hockey League between 1967 and 2022.  They performed a matched cohort study that found no significant difference in the mortality rates of enforcers versus non-enforcers (where “enforcer” is defined based on number of career fights or average career penalty minutes).  They then perform an additional sub-analysis conditioning on players that were currently deceased and looking at just this subset of players, they find a significant difference in the mean age at death between the enforcers and non-enforcers groups (45.2 vs 55.2; p-value= 0.02) thus reaching the conclusion that enforcers died an average of 10 years earlier than non-enforcers. However, some of the statistical methods used to reach their conclusions were unconvincing, specifically the approach used to reach this last claim.  Therefore, we undertook a re-analysis of their study using a survival analysis approach, which we believe is the natural method to use to study this question.  In addition, we augmented their data by scraping hockey-reference to find more instances of players who were deceased.  Then, using a Cox proportional hazards model, we found no significant difference in survival rate between enforcers and non-enforcers when an enforcer was defined based on the number of career fights.  However, when an enforcer was defined by average career penalty minutes, a significant difference was observed, which was notably not found in the original article.  Specifically, we find an estimated hazard ratio of enforcers to non-enforcers of 2.29 (p-value: 0.00169).  Therefore, we agree with the sentiment of the original paper that there is some difference between survival of these two groups, but we believe our approach to estimating this effect is more appropriate and that the magnitude of the effect in the original article is overstated.
  \vspace{2mm}
  | Keywords: survival analysis, hockey
bibliography: references.bib
fontsize: 10pt
link-citations: true
linkcolor: cyan
urlcolor: cyan
output:
  pdf_document:
    df_print: kable
    number_sections: true
    keep_tex: true
header-includes:
 \usepackage{setspace}
 \setstretch{1}
 \usepackage{float}
 \floatplacement{figure}{t}
---

@@

However, we believe that some of the methods used in @Popkin2023 to reach these conclusions are not statistically appropriates.  Most notably, the method used to reached their most sensational finding - enforcers died on average about 10 years earlier than their non-enforcer counterparts - is statistically unsuitable. Specifically, to reach this conclusion, they use the data from the matched cohort study but then condition on death.  This effectively breaks the matching that they appropriately used to compare mortality rates, and therefore this conclusion is not statistically valid.

@@

# Data 
## Data from orignal paper
We obtained the two data sets used in @Popkin2023: general table (n = 7,432) and death table (n = 45).  As in @Popkin2023, we considered only players who's first season was 1967 or later and played any games between the 1967-68 and the 2021-22 NHL seasons.  This leaves 6,039 players after filtering (the exact number mentioned in @Popkin2023).  We then removed 4 players with missing height, leaving us with `r nrow(dat)` (6,035) remaining observations, of which 252 are deaths, for our re-analysis.

@@

We note than in @Popkin2023, they find 183 E-P enforcers; we have only 182 in our data set (check that this is because of the missing height issue.)

## Additional Data
While @Popkin2023 only looks at a total of 45 deceased players  (21 enforcers and 24 controls), we identified `r sum(dat$dead)` players who were included in the data used in @Popkin2023 who were deceased as of May 7, 2023 based on data from hockey-reference.com.

@@

<!-- The same definitions of enforceres presented in @Popkin2023 are used this current manuscript for comparison purposes.   -->

<!-- (Note: In our analysis I only have 182 players who are defined as EP) -->

<!-- add a table here with summary statistics comparing the groups EF and EP.-->

@@

\caption{Comparing enforcers to non enforcers when enforcer is defined as E-F \label{tab:EDAEP}}
\end{table}

@@

# Methods {#sec:methods}
The analysis presented here utilizes all hockey players in the data, both alive and deceased, in a survival analysis to explore differences in survival times between enforcers an non-enforcers, which we believe to be a more appropriate method for analyizing the problem in this setting.  Rather than focusing on mortality rates, as in @popkin2023, we instead focus on survival times and leverage survival analysis techniques, specifically Cox proportional hazards models, to look for differences in survival times of enforcers versus non-enforcers.

>>>>

---
title: |
  |  \large Enforcers are not okay, probably.  
author: | 
abstract: |
 | A recent study looked at the relationship between long-term mortality and being an enforcer in the National Hockey League between 1967 and 2022.  They performed a matched cohort study that found no significant difference in the mortality rates of enforcers versus non-enforcers (where “enforcer” is defined based on number of career fights or average career penalty minutes).  They then perform an additional sub-analysis conditioning on players that were currently deceased and looking at just this subset of players, they find a significant difference in the mean age at death between the enforcers and non-enforcers groups (45.2 vs 55.2; p-value= 0.02) thus reaching the conclusion that enforcers died an average of 10 years earlier than non-enforcers. However, some of the statistical methods used to reach their conclusions were unconvincing, specifically the approach used to reach this last claim.  Therefore, we undertook a re-analysis of their study using a survival analysis approach, which we believe is the natural method to use to study this question.  In addition, we augmented their data by scraping hockey-reference to find more instances of players who were deceased.  Then, using a Cox proportional hazards model, we found no significant difference in survival rate between enforcers and non-enforcers when an enforcer was defined based on the number of career fights.  However, when an enforcer was defined by average career penalty minutes, a significant difference was observed, which was notably not found in the original article.  Specifically, we find an estimated hazard ratio of enforcers to non-enforcers of 2.29 (p-value: 0.00169).  Therefore, we agree with the sentiment of the original paper that there is some difference between survival of these two groups, but we believe our approach to estimating this effect is more appropriate and that the magnitude of the effect in the original article is overstated.
  \vspace{2mm}
  | Keywords: survival analysis, hockey
bibliography: references.bib
fontsize: 10pt
link-citations: true
linkcolor: cyan
urlcolor: cyan
output:
  pdf_document:
    df_print: kable
    number_sections: true
    keep_tex: true
header-includes:
 \usepackage{setspace}
 \setstretch{1}
 \usepackage{float}
 \floatplacement{figure}{t}
---

@@

However, we believe that some of the methods used in @Popkin2023 to reach these conclusions are not statistically appropriate.  Most notably, the method used to reached their most sensational finding - enforcers died on average about 10 years younger than their non-enforcer counterparts - is statistically unsuitable. Specifically, to reach this conclusion, they use the data from the matched cohort study but then condition on death.  This effectively breaks the matching that they appropriately used to compare mortality rates, and therefore this conclusion is not statistically valid.

@@

# Data 
## Data from orignal paper
We obtained the two data sets used in @Popkin2023: general table (n = 7,432) and death table (n = 45).  As in @Popkin2023, we considered only players who's first season began in 1967 or later and played any games between the 1967-68 and the 2021-22 NHL seasons.  This leaves 6,039 players after filtering (the exact number mentioned in @Popkin2023). We then removed 4 players with missing height, leaving us with `r nrow(dat)`  remaining observations, for our re-analysis.

@@

## Additional Data
While @Popkin2023 only looks at a total of 45 deceased players  (21 enforcers and 24 controls), we identified `r sum(dat$dead)` players who were included in the data used in @Popkin2023 who were deceased as of May 7, 2023 based on data from hockey-reference.com. Of these additional `r sum(dat$dead) - nrow(death_JAMA)` observations of death, 2 fall in the E-F cohort, and 4 fall in the E-P cohort.

@@

@@

\caption{Comparing enforcers to non enforcers when enforcer is defined as E-P \label{tab:EDAEP}}
\end{table}

@@

# Methods {#sec:methods}
The analysis presented here utilizes all hockey players in the data, both alive and deceased, in a survival analysis to explore differences in survival times between enforcers an non-enforcers, which we believe to be a more appropriate method for analyizing the problem in this setting.  Rather than focusing on mortality rates, as in @Popkin2023, we instead focus on survival times and leverage survival analysis techniques, specifically Cox proportional hazards models, to look for differences in survival times of enforcers versus non-enforcers.
